Speed and Precision in Range Analysis

### 2012-04-18 doc/Manuscript/SBLP/paper.tex
Fixed my email again.
Changed the tex, and a the figure about context sensitiveness.

\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
\email{\{victorsc,raphael,igor.rafael,fernando\}@dcc.ufmg.br}
}

@@

% Summary of our experiments
We have implemented an industrial-quality range analysis algorithm in the
LLVM compiler~\cite{Lattner04}.
As we show in Section~\ref{sub:showdown}, our implementation, which is
publicly available, is able to analyze programs with over one million assembly
instructions in ten seconds.
And our implementation is not a straw-man: it produces very precise results.
We have compared the results produced by our implementation with the results
obtained via a dynamic profiler, which we have also implemented.
As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
benchmarks we are able to estimate tight ranges for almost half of all the
integer variables present in these programs.

@@

As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
analysis algorithm has many optional modules.
These modules provide the user with the possibility to choose between
more precise results, or a faster analysis.
Given the number of options, the design space of a range analysis algorithm
is vast.
In this section we try to cover some of the most important tradeoffs.
Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
benchmark suite, precision versus speed for different configurations of
our implementation.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variables.
It is very important to notice that we do not consider constants in our statistics of precision.
In other words, we only measure bitwidth reduction in variables that a constant
propagation step could not remove.

@@

The greatest source of improvement in our implementation is the use of strongly
connected components.
In order to propagate ranges across the constraint graph, we fragment it
into strongly connected components, collapse each of these components into
single nodes, and sort the resulting directed acyclic graph topologically.
We then solve the range analysis problem for each component individually.
Once we have solved a component, we propagate its ranges to the next
components, and repeat the process until we walk over the entire constraint
graph.
It is well-known that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}.
In our case, the results are dramatic, mostly in terms of speed, but also in
terms of precision.
Figure~\ref{fig:impactSCC} shows the speed up that we gain by using strong
components.
We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
In some cases, as in \texttt{xalancbmk} the analysis on strong components is
450x faster.

@@

The strong components improve the precision of our growth analysis.
As we see in Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
it is 40\% more precise.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$0, +\infty$.

@@

A naive implementation of range analysis would be intra-procedural; that is,
would solve the range analysis problem once for each function.
However, we can gain in precision by performing it inter-procedurally.
An inter-procedural implementation allows the results found for a function $f$
to flow into other functions that $f$ calls.
Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
program seen in Figure~\ref{fig:overall_view}(a).
The trivial way to produce an inter-procedural implementation is to insert
into the constraint system assignments from the actual parameter names to the
formal parameter names.
In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
of information from $0$, the actual parameter, to $k_0$, the formal parameter
of function \texttt{foo}.

@@

Figure~\ref{fig:wholeImpact} shows that the whole program analysis increases
moderately the precision of our analysis.
We show results for the five largest programs in three different categories of
benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointingly: on the average for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of the total program bitwidth.
By using the inter-procedural version we can increase this number to 8.89\%.
A manual inspection of the SPEC programs reveal that this result is expected:
they manipulate files, and their texts do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is brought down by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results.

@@

Another way to increase the precision of range analysis is by using a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
An usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
Bars are normalized to the time to run the intra-procedural analysis
without inlining.
On the average, the intra-procedural mode is 28.92\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is larger: 45.87\%.
These numbers are close because we run our analysis on strong components.

@@

If strong components account for the largest gains in speed, the choice of
a suitable program representation is responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our speed, because it increases the
size of the target program.
We have tried our analysis in two different program representations: the
Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
Single Assignment (e-SSA) form~\cite{Bodik00}.
The SSA form gives us a faster, albeit more imprecise, analysis.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that created variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases because the e-SSA form programs are larger than the SSA
form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
However, this growth is small: in none of the integer programs in SPEC
CPU 2006 we verified an increase in code size of more than 9\%.
But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
are remarkable, as we show in Figure~\ref{fig:precESSA}.

@@

We have implemented the widening operator used in the growth analysis in
two different ways.
The first way, which we call {\em simple}, is based on Cousot and
Cousot's original widening operator~\cite{Cousot77}.
This operator is shown in Figure~\ref{fig:widening}, and it is the one
used in Figure~\ref{fig:ex_partition_grow_crop}(b).
The second widening strategy, which we call {\em jump-set widening} consists
in using the constants that appear in the program text, in sorted order, as
the next limits of each interval after widening is applied.
This operator is common in implementations of range
analysis~\cite[p.228]{Nielson99}.
There are situations in which jump-set widening produces better results
than the simple operator.
Figure~\ref{fig:jumpSet} shows an example taken from the code of
SPEC CPU \texttt{bzip2}.
Part of the constraint graph of the program in Figure~\ref{fig:jumpSet}(a)
is given in Figure~\ref{fig:jumpSet}(b).
The result of applying the simple operator is shown in
Figure~\ref{fig:jumpSet}(c).
Jump-set widening would use the lattice in Figure~\ref{fig:jumpSet}(d),
instead of the lattice in Figure~\ref{fig:widening}(Right).
This lattice would yield the result given in Figure~\ref{fig:jumpSet}(e),
which is more precise.

@@

Another way to improve the precision of growth analysis is to perform a few
rounds of abstract interpretation on the constraint graph, and, in case the
process does not reach a fix point, only then applying the widening operator.
Each round consists in evaluating all the constraints, and updating the
intervals that change from one evaluation to the other.
The simple widening strategy, coupled with one round of abstract interpretation
would give us precise results in the example from Figure~\ref{fig:jumpSet}.
We have experimented with 0 and 16 iterations before doing widening, and the
overall result, for the programs in the SPEC CPU 2006 suite is given in
Figure~\ref{fig:space}.
Figure~\ref{fig:wideningPrec} shows some of these results in more details
for a broader set of benchmarks.
In general jump-set widening improves the precision of our results in
non-trivial ways.
Nevertheless, the simple widening operator preceded by 16 rounds of
abstract interpretation in general is more precise than jump-set widening
without any cycle of pre-evaluation, as we see in Figure~\ref{fig:space}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/wideningPrec}
\end{center}
\caption{\label{fig:wideningPrec}
The impact of the widening strategy on precision.
Each bar gives precision in \%bitwidth reduction.}
\end{figure}

>>>>

\address{UFMG -- 6627 Ant\^{o}nio Carlos Av, 31.270-010, Belo Horizonte, Brazil
\email{\{victorsc,raphael,igor,fernando\}@dcc.ufmg.br}
}

@@

% Summary of our experiments
We have implemented an industrial-quality range analysis algorithm in the
LLVM compiler~\cite{Lattner04}.
As we show in Section~\ref{sub:showdown}, our implementation, which is
publicly available, is able to analyze programs with over one million assembly
instructions in sixteen seconds.
And our implementation is not a straw-man: it produces very precise results.
We have compared the results produced by our implementation with the results
obtained via a dynamic profiler, which we have also implemented.
As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
benchmarks we are able to estimate tight ranges for almost half of all the
integer variables present in these programs.

@@

As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
analysis algorithm has many optional modules.
These modules give the user the chance to choose between more precise results,
or a faster analysis.
Given the number of options, the design space of a range analysis algorithm
is vast.
In this section we try to cover some of the most important tradeoffs.
Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
benchmark suite, precision versus speed for different configurations of
our implementation.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variable.
It is very important to notice that we do not consider constants in our statistics of precision.
In other words, we only measure bitwidth reduction in variables that a constant
propagation step could not remove.

@@

The greatest source of improvement in our implementation is the use of strongly
connected components.
In order to propagate ranges across the constraint graph, we fragment it
into strongly connected components, collapse each of these components into
single nodes, and sort the resulting directed acyclic graph topologically.
We then solve the range analysis problem for each component individually.
Once we have solved a component, we propagate its ranges to the next
components, and repeat the process until we walk over the entire constraint
graph.
It is well-known that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}.
In our case, the results are dramatic, mostly in terms of speed, but also in
terms of precision.
Figure~\ref{fig:impactSCC} shows the speedup that we gain by using strong
components.
We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
In some cases, as in \texttt{xalancbmk} the analysis on strong components is
450x faster.

@@

The strong components improve the precision of our growth analysis.
According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
strong components increase our precision by 40\%.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$0, +\infty$.

If strong components account for the largest gains in speed, the choice of
a suitable program representation is responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our speed, because it increases the
size of the target program.
We have tried our analysis in two different program representations: the
Static Single Assignment (SSA) Form~\cite{Cytron91}, and the Extended Static
Single Assignment (e-SSA) form~\cite{Bodik00}.
The SSA form gives us a faster, albeit more imprecise, analysis.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that creates variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases because the e-SSA form programs are larger than the SSA
form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
However, this growth is small: in none of the integer programs in SPEC
CPU 2006 we verified an increase in code size of more than 9\%.
But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
are remarkable, as seen in Figure~\ref{fig:precESSA}.
These gains happen because the e-SSA format lets the analysis to use the
results of comparisons to narrow the ranges of variables.

@@

A naive implementation of range analysis would be intra-procedural; that is,
would solve the range analysis problem once per each function.
However, we can gain in precision by performing it inter-procedurally.
An inter-procedural implementation allows the results found for a function $f$
to flow into other functions that $f$ calls.
Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
program seen in Figure~\ref{fig:overall_view}(a).
The trivial way to produce an inter-procedural implementation is to insert
into the constraint system assignments from the actual parameter names to the
formal parameter names.
In our example of Figure~\ref{fig:intra}, our constraint graph contains a flow
of information from $0$, the actual parameter, to $k_0$, the formal parameter
of function \texttt{foo}.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointingly: on the average for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs manipulate files, and their source codes do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

@@

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
An usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
If a function is called more than once, then its formal parameters will receive
information from many actual parameters.
We use $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
Bars are normalized to the time to run the intra-procedural analysis
without inlining.
On the average, the intra-procedural mode is 28.92\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is 45.87\%.
These numbers are close because our runtime is bound to the
size of the strong components.
We have observed that function inlining does not increase too much these
components.

@@

@@

We have implemented the widening operator used in the growth analysis in
two different ways.
The first way, which we call {\em simple}, is based on Cousot and
Cousot's original widening operator~\cite{Cousot77}.
This operator is shown in Figure~\ref{fig:widening}, and it is the one
used in Figure~\ref{fig:ex_partition_grow_crop}(b).
The second widening strategy, which we call {\em jump-set widening} consists
in using the constants that appear in the program text, in sorted order, as
the next limits of each interval after widening is applied.
This operator is common in implementations of range
analysis~\cite[p.228]{Nielson99}.
There are situations in which jump-set widening produces better results
than the simple operator.
Figure~\ref{fig:jumpSet} shows an example taken from the code of
SPEC CPU \texttt{bzip2}.
Part of the constraint graph of the program in Figure~\ref{fig:jumpSet}(a)
is given in Figure~\ref{fig:jumpSet}(b).
The result of applying the simple operator is shown in
Figure~\ref{fig:jumpSet}(c).
Jump-set widening would use the lattice in Figure~\ref{fig:jumpSet}(d),
instead of the lattice in Figure~\ref{fig:widening}(Right).
This lattice yields the result given in Figure~\ref{fig:jumpSet}(e),
which is more precise.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Another way to improve the precision of growth analysis is to perform a few
rounds of abstract interpretation on the constraint graph, and, in case the
process does not reach a fix point, only then to apply the widening operator.
Each round of abstract interpretation consists in evaluating all the
constraints, and then updating the intervals that change from one evaluation to
the other.
For instance, in Figure~\ref{fig:jumpSet} one round of abstract interpretation,
coupled with the simple widening operator, would be enough to reach the
fix point of that constraint system.
We have experimented with 0 and 16 iterations before doing widening, and the
overall result, for the programs in the SPEC CPU 2006 suite is given in
Figure~\ref{fig:space}.
Figure~\ref{fig:wideningPrec} shows some of these results in more details
for the five largest benchmarks in this collection.
In general jump-set widening improves the precision of our results in
non-trivial ways.
Nevertheless, the simple widening operator preceded by 16 rounds of
abstract interpretation in general is more precise than jump-set widening
without any cycle of pre-evaluation, as we see in Figure~\ref{fig:wideningPrec}.

\begin{figure}[t!]
\begin{small}
\begin{center}
\renewcommand{\tabcolsep}{0.2cm}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|} \hline
Benchmark &    Size & 0 + Simple &   16 + Simple &     0 + Jump & 16 + Jump \\ \hline
hmmer     &  38,409 &       9.98 & 11.40 (12.45) & 10.98 (9.11) & 11.40 (12.45) \\ \hline
gobmk     &  84,846 &       8.15 &  9.93 (17.92) &  9.02 (9.64) & 10.13 (19.54) \\ \hline
h264ref   &  97,494 &      12.58 &  13.11 (4.04) & 13.00 (3.23) & 13.11 (4.04) \\ \hline
xalancbmk & 352,423 &       7.71 &   7.98 (3.38) &  7.95 (3.02) & 7.98 (3.38) \\ \hline
gcc       & 449,442 &      16.09 &  16.63 (3.25) & 16.41 (1.95) & 16.64 (3.31) \\ \hline
\end{tabular}
\end{center}
\end{small}
\caption{\label{fig:wideningPrec}
Variation in the precision of our analysis given the widening strategy.
The size of each benchmark is given in number of variable nodes in the
constraint graph.
Precision is given in percentage of bitwidth reduction.
Numbers in parenthesis are percentage of gain over 0 + Simple.}
\end{figure}
